---
title: "NFL Punt Rule Change Win Probability Impact"
output: html_notebook
bibliography: references.bib
---

This report examines the impact on a team's win probability from our proposed
rule changes to incentivize less returns on punts by NFL teams.

We first gather NFL.com play-by-play data accessed via the `nflscrapR` in `R`, 
with all pre-, post-, and regular season data for the 2016 and 2017 seasons available here: https://github.com/ryurko/nflscrapR-data/tree/master/play_by_play_data

```{r, warning = FALSE, message = FALSE}
# Access tidyverse:
# install.packages("tidyverse")
library(tidyverse)

# Access nflscrapR:
# install_packages("devtools")
# devtools::install_github(repo = "maksimhorowitz/nflscrapR")
library(nflscrapR)

# Load the play-by-play data from the 2016 and 2017 season available in the repository above:

# Create the combination of year and type of games to access data for:
season_type_data <- expand.grid(c(2016, 2017), 
                              c("pre", "regular", "post")) %>%
  rename(season = Var1, type = Var2)

# Join all of these seasons together:
pbp_data <- map2_dfr(season_type_data$season, as.character(season_type_data$type),
                    .f = function(season, type) {
                      # Shorten abbreviation to reg for regular season:
                      type_abbr <- if_else(type == "regular", "reg", type)
                      
                      # Now load the data:
                      read_csv(paste0(
                        "https://raw.githubusercontent.com/ryurko/nflscrapR-data/master/play_by_play_data/", 
                        type, "_season/", type_abbr, "_pbp_", season, ".csv"),
                        progress = FALSE)
                    })

# Access all punts as well as any play following the punt in order to have
# proper win probability calculation:
punt_pbp_data <- pbp_data %>%
  group_by(game_id) %>%
  # Create booleans for the certain situations we're interested in making
  # rule changes for:
  mutate(follow_punt = ifelse(lag(play_type == "punt"), TRUE, FALSE),
         follow_fair_catch = ifelse(lag(punt_fair_catch) == 1, TRUE, FALSE),
         follow_punt_oob = ifelse(lag(punt_out_of_bounds) == 1, TRUE, FALSE),
         follow_punt_touchback = ifelse(follow_punt & lag(touchback == 1), 
                                        TRUE, FALSE),
         follow_downed_punt = ifelse(lag(punt_downed) == 1, TRUE, FALSE)) %>%
  filter(play_type == "punt" |
         follow_punt) %>%
  ungroup()

```

Since our proposed rule change to lower the concussion rate on punts is to simply
incentivize teams to return less often, it is necessary to view the impact of our
proposed incentives on the game itself. We propose the following two rule changes:

#. Return team gains 5 additional yards on a fair catch (__incentive for return team__).
#. Return team loses 5 yards from the point the ball goes out of bounds in the air (__incentive for the punting team__). 

There are no proposed incentive structures for punts that land in play, regular 
rules will apply.

The histogram below displays the distribution of return yards for all punt 
returns (excluding fair catch, touchbacks, out of bounds, and downed punts) with 
a reference line at our proposed shift of five yards. Roughly `r round(digits = 2, nrow(filter(punt_pbp_data,play_type == "punt", punt_fair_catch == 0 & punt_out_of_bounds == 0 & touchback == 0 & punt_downed == 0, return_yards <= 5)) / nrow(filter(punt_pbp_data,play_type == "punt", punt_fair_catch == 0 & punt_out_of_bounds == 0 & touchback == 0 & punt_downed == 0)) * 100)`% of punt
returns are less than or equal to 5 yards.

```{r, fig.width = 4, fig.height = 3}
punt_pbp_data %>%
  filter(play_type == "punt",
         punt_fair_catch == 0 & punt_out_of_bounds == 0 & 
           touchback == 0 &
           punt_downed == 0) %>%
  ggplot(aes(x = return_yards)) +
  geom_histogram(fill = "darkblue", color = "black",
                 breaks = seq(-15, 100, by = 5)) +
  scale_x_continuous(limits = c(-15, 100), breaks = seq(-10, 100, by = 10)) +
  geom_rug(color = "darkblue", alpha = 0.75) +
  geom_vline(xintercept = 5, color = "darkorange", linetype = "dashed", size = 1.5) +
  labs(x = "Return yards",
       y = "Frequency", 
       title = "Distribution of punt return yards from pre-, post-, and regular season games (2016-17)",
       subtitle = "Only includes punts with returns; dashed vertical line denoting five yard returns",
       caption = "Pelechrinis, Yurko, Ventura (2019)\nData courtesy of NFL.com, accessed with `nflscrapR`") +
  theme_bw() 
```

The empirical CDF below is displaying the same information, but now is it easier
to see that the proposed yardage shift of 5 yards is comparable to the median
of return yards which is `r punt_pbp_data %>% filter(play_type == "punt", punt_fair_catch == 0, punt_out_of_bounds == 0, touchback == 0, punt_downed == 0) %>% pull(return_yards) %>% median()` yards.

```{r, fig.width = 4, fig.height = 3}
punt_pbp_data %>%
  filter(play_type == "punt",
         punt_fair_catch == 0 & punt_out_of_bounds == 0 & 
           touchback == 0 & punt_downed == 0) %>%
  ggplot(aes(x = return_yards)) +
  stat_ecdf(color = "darkblue") +
  scale_x_continuous(limits = c(-15, 100), breaks = seq(-10, 100, by = 10)) +
  geom_vline(xintercept = 5, color = "darkorange", linetype = "dashed") +
  labs(x = "Return yards",
       y = "Cumulative proportion of returns", 
       title = "Empirical CDF of punt return yards from pre-, post-, and regular season games (2016-17)",
       subtitle = "Only includes punts with returns; dashed vertical line denoting five yard returns",
       caption = "Pelechrinis, Yurko, Ventura (2019)\nData courtesy of NFL.com, accessed with `nflscrapR`") +
  theme_bw() 
```

Next, using the win probability in `nflscrapR` [@Yurko18], we calculate the 
win probability added from the proposed change for all punt returns with a
fair catch or out of bounds, and compare it to the win probability
added from returns. This means we are comparing the change in win probability
from where the returning team would have hypothetically received the ball given
the current rules to where the return ends or the 5 yard shift from the proposed
rule change. Downed punts and touchbacks are not included in the figure below.

```{r, warning = FALSE, message = FALSE, fig.width=5, fig.height=5, fig.align="center"}

# Create a new dataset of punts and following plays by adjusting the following
# plays starting field position based on the proposed rule change for the
# particular situation:
punt_adj_data <- punt_pbp_data %>%
  # Create a label for all punts and following plays based on whether it was
  # a normal return, fair catch, out of bounds, or touchback:
  mutate(punt_type = case_when(
    (play_type == "punt" & punt_fair_catch == 1) |
      follow_fair_catch ~ "fair_catch",
    (play_type == "punt" & punt_out_of_bounds == 1) |
      follow_punt_oob ~ "out_of_bounds",
    (play_type == "punt" & touchback == 1) |
      follow_punt_touchback ~ "touchback",
    (play_type == "punt" & punt_downed == 1) |
      follow_downed_punt ~ "downed",
    TRUE ~ "return"),
    # Create a copy variable for the map function to use easily:
    punt_type_copy = punt_type,
    # Create a variable that has the yard adjustment with respect to the return team:
        yard_adjust = case_when(
        punt_type == "fair_catch" ~ 5,
        punt_type %in% c("out_of_bounds", "touchback") ~ -5,
        TRUE ~ -1 * as.numeric(lag(return_yards)))) %>%
  # Remove downed and touchback punt rows:
  filter(!(punt_type %in% c("downed", "touchback"))) %>%
  # Now group by this punt type:
  group_by(punt_type) %>%
  # Nest this dataframe:
  nest() %>%
  # Now apply a function to each of the datasets that calculates the change 
  # in win probability based on the proposed rule change:
  mutate(data = map(data,
                    function(x) {
                      x %>% 
                        # Create an adjusted yardline_100 based on the yard_adjust,
                        # capping at 99 yards away from opponents end zone:
                        mutate(adj_yardline_100 = pmin(yardline_100 - yard_adjust, 99),
                               less_than_10 = ifelse(yardline_100 <= 10, TRUE, FALSE),
                               adj_yardline_100 = ifelse(less_than_10, ceiling(yardline_100 / 2),
                                                         adj_yardline_100)) %>%
                        # Select the necessary columns for win probability calculations and other info:
                        dplyr::select(game_id, play_type, punt_fair_catch, 
                                      follow_punt, follow_fair_catch,
                                      posteam, defteam, desc, yardline_100, 
                                      adj_yardline_100, half_seconds_remaining, 
                                      game_seconds_remaining, down, ydstogo, 
                                      goal_to_go, qtr, score_differential,
                                      posteam_timeouts_remaining, 
                                      defteam_timeouts_remaining,
                                      return_yards,
                                      # Including the original WP, EP, and WPA:
                                      wp, ep, wpa, yard_adjust, punt_type_copy) %>%
                        # Rename the original WP and EP columns
                        rename(original_wp = wp,
                               original_EP = ep) %>%
                        # Now calculate the expected points for the adjusted yard line:
                        calculate_expected_points("half_seconds_remaining",
                                                  "adj_yardline_100", "down", 
                                                  "ydstogo", "goal_to_go") %>%
                        calculate_win_probability("half_seconds_remaining", 
                                                  "game_seconds_remaining",
                                                  "score_differential", "qtr", 
                                                  "posteam_timeouts_remaining",
                                                  "defteam_timeouts_remaining", 
                                                  "ep") %>%
                        # Calculate the win probability added for rule change:
                        group_by(game_id) %>%
                        # If the type if just a return, then take the difference
                        # of the original WP with the adjusted from where the ball
                        # was returned from (with respect to the return team) 
                        # otherwise calculate rule change WPA:
                        mutate(rule_change_wpa = if_else(punt_type_copy == "return",
                                                         original_wp - wp,
                                                         wp - original_wp)) %>%
                        ungroup()
                    })) %>%
  # Unnest the data and ungroup:
  unnest() %>% ungroup()

# Save this file for the kernel to access:
# write.csv(punt_adj_data, "~/Documents/nflscrapR-data/R/punt_competition/punt_adj_data.csv")


# Visualize the distributions of WPA due from returns compared to the proposed
# rule changes using histograms:
punt_adj_data %>%
  # Only look at the plays following punts, removing QB kneels and overtime,
  # as well as touchbacks:
  filter(follow_punt, play_type != "qb_kneel", qtr != 5) %>%
  # Reorder and relabel the punt_type:
  mutate(punt_type = fct_relevel(punt_type, "return", "fair_catch", 
                                 "out_of_bounds"),
         punt_type = fct_recode(punt_type, `Historical punt returns` = "return", 
                                              `Rule change #1: 5 yard fair catch adjustment` = "fair_catch",
                                              #`Rule change: touchback` = "touchback",
                                              `Rule change #2: -5 yards out of bounds adjustment` = "out_of_bounds")) %>%
  # Create beeswarm plots with boxplots overtop:
  ggplot(aes(x = rule_change_wpa)) +
  geom_histogram(aes(fill = punt_type), color = "black",
                 breaks = seq(-.05, .35, by = .005)) +
  ggthemes::scale_fill_colorblind(guide = FALSE) +
  facet_wrap(~punt_type, ncol = 1) +
  # geom_density_ridges(aes(y = punt_type), rel_min_height = 0.01, jittered_points = TRUE, color = "white",
  #                     position = position_points_jitter(width = 0, height = 0),
  #                     point_shape = '|', point_size = 1, point_alpha = 0.7, alpha = 0.8,
  #                     fill = "darkblue", point_color = "darkblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 1.5) +
  geom_text(data = data.frame("punt_type" = c("Rule change #1: 5 yard fair catch adjustment",
                                              "Rule change #2: -5 yards out of bounds adjustment"),
                              "text_label" = c("Rule change #1\nfavors return team",
                                               "Rule change #2\nfavors punting team"),
                              "x" = c(.06, .06),
                              "y" = c(450, 450)), aes(x, y, label = text_label),
            color = "darkred", size = 6) +
  scale_x_continuous(limits = c(-0.05, .2), breaks = seq(-.05, .2, by = .05)) +
  labs(title = "Distribution of win probability added from returns compared to proposed rule changes",
       #subtitle = "Calculated with respect to possession team",
       subtitle = "All punt returns from pre-, post-, and regular season games in 2016 and 2017",
       caption = "Pelechrinis, Yurko, Ventura (2019)\nData courtesy of NFL.com, accessed with `nflscrapR`",
       x = "Win probability added (WPA) with respect to return team",
       y = "Frequency") +
  #scale_y_discrete(expand = c(0.01, 0.01)) +
  #scale_x_continuous(limits = c(-.5, .5)) +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 14),
        plot.title = element_text(size = 16),
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(size = 12),
        strip.background = element_blank(),
        strip.text = element_text(size = 14))  


```

The figure below just displays the empirical CDF form of the above chart:


```{r, warning = FALSE, message = FALSE,  fig.align="center"}
ecdf_war_plot <- punt_adj_data %>%
  # Only look at the plays following punts, removing QB kneels and overtime:
  filter(follow_punt, play_type != "qb_kneel", qtr != 5) %>%
  # Reorder and relabel the punt_type:
  mutate(punt_type = fct_relevel(punt_type, "return", "fair_catch",
                                 "out_of_bounds"),
         punt_type = fct_recode(punt_type, `Historical punt returns` = "return", 
                                              `Rule change #1: 5 yard fair catch adjustment` = "fair_catch",
                                              #`Rule change: touchback` = "touchback",
                                              `Rule change #2: -5 yards out of bounds adjustment` = "out_of_bounds")) %>%
  # Create beeswarm plots with boxplots overtop:
  ggplot(aes(x = rule_change_wpa, color = punt_type)) +
  stat_ecdf(alpha = 0.8) +
  ggthemes::scale_color_colorblind() +
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 1.5) +
  labs(title = "Empirical CDF of win probability added from returns compared to proposed rule changes",
       #subtitle = "Calculated with respect to possession team",
       subtitle = "All punt returns from pre-, post-, and regular season games in 2016 and 2017",
       caption = "Data courtesy of NFL.com, WPA generated w/ `nflscrapR`",
       x = "Win probability added (WPA) with respect to return team",
       y = "Cumulative proportion of punt return type", 
       color = "Punt return type") +
  #scale_y_discrete(expand = c(0.01, 0.01)) +
  #scale_x_continuous(limits = c(-.5, .5)) +
  theme_bw() +
  theme(axis.title = element_text(size = 10),
        axis.text = element_text(size = 10),
        plot.title = element_text(size = 12),
        plot.subtitle = element_text(size = 10),
        plot.caption = element_text(size = 8),
        strip.background = element_blank(),
        strip.text = element_text(size = 14),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10))  
library(cowplot)
ecdf_legend <- get_legend(ecdf_war_plot)

plot_grid(ecdf_war_plot + theme(legend.position = "none"),
          ecdf_legend, ncol = 1, rel_heights = c(3,1))

```


# References
